 \chapter*{Acknowledgements}
\label{C:Acknowledgements}
\addcontentsline{toc}{section}{\nameref{C:Acknowledgements}}

% Linda
First and foremost, I want to thank my supervisor Linda Cremonesi for her constant support throughout my PhD. Linda taught me many things, but most importantly how to be an independent researcher. Thank you for these four years full of great fun and great physics. This thesis is only possible thanks to her dedication.

% Stefano, Claire and the QMUL neutrino group
Thank you to Stefano Moretti and Claire Shepherd-Themistocleous, my supervisors at Southampton and RAL. They have provided me with invaluable help during this adventure. I would also like to thank the other members of the QMUL neutrino group -- Abbey Waldron, Alex Booth, Jon Hays, Nicola McConkey, Prabhjot Singh and Simranjit Singh Chhibra -- for their guidance. Our Liquid Argon weekly meetings have always been the perfect space to discuss physics.

% QMUL PhD students
I came to the conclusion that the worth of a PhD increases exponentially with the number of friends you make along the way. That is why I thank my fellow students in the QMUL particle physics group: Nathan, Aled, Kevin, Sid, Zahaab, Akeem, Callum, James, Talha, Albanik, Arnav, Jocelyn, Lottie, and Oscar. I could not imagine my life in this university without them. To more trips to The Cherry, karaoke parties and pub quizzes.

% DUNE DAQ FWTPG crew
I also want to acknowledge the work of the DUNE DAQ Firmware Readout group. In particular, I would like to thank Alessandro Thea and Ivana Hristova for their help and the long chats about the DAQ. From my time on LTA at CERN, I also want to say thank you to Shyam Bhuller and Klaudia Wawrowska. All of them made going to the office in Pr\'{e}vessin really worth it.

% ND-GAr group
For the last two years, ND-GAr has been main research focus. I want to thank the conveners of the group -- Alysia Marino, Patrick Dunne and Tanaz Mohayai -- for the opportunity to contribute to their efforts. I also want to express my gratitude to Alan Bross, Leo Bellantoni, Tom Junk and Vivek Jain, for their insight and all the chats we have had, from physics to hardware and software. And of course, thank you to the other PhD students in the group: Callum Cox, Hank Hua, and Naseem Khan.

% The rest of DUNE
I am so grateful to be a part of the DUNE collaboration. During my PhD I have met many collaborators who are both excellent scientists and amazing people. Particularly, I want to thank Alex Wilkinson and Robert Kralik, with whom I share my excitement about neutrino physics either in conferences, collaboration meetings or in random pubs around London.

% Family and (other) friends
I want to thank everyone else who has been a part of my life while working on this thesis. Thank you to Bea for her unconditional support and for always believing in me. Thank you to Ana for her company in these last months of writing. Thank you to Giovanna for being there in the rough spots of this journey, especially at the beginning. Thank you to my friends in Spain for their patience and understanding. Finally, I would like to thank my parents and my siblings, who always supported and loved me. This thesis is also yours.